TacticToe: Learning to Prove with Tactics

### 2018-02-19 tactictoe_jv.tex
example

%\title{TacticToe: A theorem prover trained from formal proofs}
%\title{TacticToe: Smart guidance in tactical proof search}
%\title{TacticToe: Proof  by Tactic Selection}
\title{TacticToe: Learning the Game of Tactics}
%\title{TacticToe: Automatic tactic-level proof generation}

@@

\todo{JU review: Still feels more like an intro than an abstract. Could be
shorter and punchier.}
\begin{abstract}
Formalization of mathematics in interactive theorem provers (ITPs) relies today on many proof automation methods.
Some of these methods are general-purpose and with enough time they can prove any provable conjecture.
Many formalized proof scripts however also take advantage of specialized rules and theory-based strategies.
In the HOL4 ITP system, practically all of them are implemented as \emph{tactics}.

@@

\tableofcontents
\todoi{Comparison with hammers in usability, easier reconstruction nice proofs}

\section{Introduction}

Many of the state-of-the-art interactive theorem provers (ITPs) such as
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight},
  \isabelle~\cite{isabelle}
  and \coq~\cite{coq-book} provide high-level parameterizable tactics for constructing proofs.
  Such tactics typically analyze the current proof state (goal and
  assumptions), then apply non-trivial proof transformations that get
  expanded into many basic kernel-level inferences or large parts of a proof term.
%\marginpar{not Coq}
  In this work we develop a tactic-level automation procedure for the \holfour ITP
  which guides the selection of tactics by learning from previous
  proofs.  Instead of relying on translation to first-order automated
  theorem provers (ATPs) as done by the hammer
  systems~\cite{hammers4qed,tgck-cpp15}, our technique
%\marginpar{too many cites?}
%BlanchetteGKKU16,holyhammer,mizAR40
  searches directly for sequences of tactic applications that lead to
  an ITP proof, thus avoiding the translation and proof-reconstruction phases needed by the hammers.

\todoi{Rewrite the plan}
  To do this, we \emph{extract and record} tactic invocations from the ITP
  proofs (Section~\ref{sec:recording}) and
  \emph{build efficient machine learning classifiers} based on such training
  examples (Section~\ref{s:prediction}).  The learned data serves as
  guidance for a \emph{Monte Carlo tree search} that explores different
  proof paths (Section~\ref{sec:proofsearch}). The result, if
  successful, is a verified human-level proof composed of \holfour
  tactics.  The system is evaluated on a large set of theorems originating
  from \holfour (Section~\ref{s:experiments}).

@@

\paragraph{Tactics in \holfour.}
Considering concrete implementation details now, a \emph{tactic} is a function
that takes a goal (or proof obligation) and returns a list of goals (subgoals of the original goal) together
with a validation function.
Calling the validation function on a list of \emph{theorems}, corresponding to the list of subgoals, results in a theorem corresponding to the original goal.
For example, if the list of subgoals is empty, then calling the validation function on the empty list yields the originally goal sequent as a theorem.
In this way, tactics implement the plumbing of backwards, goal-directed, proof 
construction. 
Because validation functions are only important to check the final proof, we 
omit them during the description of proof search algorithm. And we note $t(g)$ 
the lists of goals produced by the application of the tactic $t$ to the goal 
$g$.

\subsection{Contributions}
This paper extends the work described in \cite{tgckju-lpar17}.
There we proposed the idea of ... and achieved ...
The additional contributions and their effect on our system are:

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern
matching constructions, opening of submodules are supported. The different
recording steps (parsing, unfolding and replaying) are regrouped into a single
\sml function.
\item Monte Carlo tree search~\cite{montecarlo} (MCTS) replaces A* as our
search algorithm. The MCTS algorithm gives a more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by predictors for
three kinds of objects: tactics, theorems and lists of goals.
\item The orthogonalization process that eliminates redundant tactics is 
described in a precise manner.
\item A tactic abstraction mechanism is introduced. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover to help \tactictoe during proof search.
\item First evaluation of \tactictoe success rate relative to
the length of the original human proofs.
\item Minimization and prettification of the generated proof improves user 
interaction.
\end{itemize}

@@

\begin{definition}\label{def:prod} (Productive tactic)
The application of a tactic on an open goal is called productive if and only if
all the following conditions are satisfied:
\begin{itemize}
\item It does not fail or timeout. To prevent tactics from looping, we
interrupt tactics after a short amount of time (0.05 seconds).
\item It does not loop, i.e its output does not contain any goals that appears 
in the ancestor of the node of the current goal.
\item It is not a parallel step, i.e its output is not a superset
of a list of goals produced from the same goal.
\end{itemize}

@@

\section{Proof Recording}\label{sec:recording}
Recording proofs in an LCF-style proof assistant can be done at different
levels.
In \holfour all existing approaches relied on modifying the kernel. This was
used
either to export the primitive inference
steps~\cite{Wong95recordingand,DBLP:conf/itp/KumarH12}
or to record dependencies of theorems~\cite{tgck-cpp15}. This was not suitable
for our
purpose of learning proving strategies at the intermediate tactic level. We
therefore
discuss recording proofs in an LCF-style proof assistant, with the focus on
\holfour
in this section.
<
Rather than relying on the underlying programming language, we parse the
original theory file containing proof scripts. This enables us to extract the
code of each tactic.
Working with the string representation of a tactic is better that working with
its value:
\begin{itemize}
\item Equality between two functions is easy to do from
their string representation. The direct consequence is that we can avoid
predicting the same tactic over and over again.
\item
The
results of
\tactictoe can be returned in a readable form to the user. In this way, the
user can learn from the feedback, analyze the proof and possibly improve on it.
An other point is that \tactictoe does not need to be installed to run
proof scripts generated by \tactictoe. In this way, the further development of
\tactictoe are not affecting the robustness of \holfour.
\item
It is difficult (probably impossible) to transfer \sml values
between theories as they are not run in the same \holfour session. In contrast,
\sml code can be exported, imported.
\item To produce a tactic value form a code is easy in \sml, which can be
achieved by using a reference to a tactic and update it with the function
\texttt{use}.
\end{itemize}

@@

In Example~\ref{ex:cs3}, we prove a theorem about lists, a domain where \tactictoe excels compared to ATPs.
Given two list $l_1$ and $l_2$ where the length of $l_1$ (noted $p$) is less
than a natural
number $n$, the theorem states updating the $n^{th}$
element of the concatenation of $l_1$ and $l_2$ is the same as
updating the $m^{th}$ element of $l_2$ where $m = n - p$. Again,
\tactictoe's proof is
much more readable starting with applying induction on $l$. It solves the base
case of the induction by rewriting and distinguish cases where $n=0$ or $n>0$
in the inductive hypothesis. It finalizes the proof with a short call to \metis
using the definition of
$LUPDATE$. Here, \tactictoe's proof is arguably smaller, faster and easier to
understand and maintain. Such proofs after an expert review could replace the
their respective original human proofs in the \holfour repository.

@@

In Example~\ref{ex:cs4}, we demonstrate the capabilities of \tactictoe on a goal that was not already proved in the \holfour libraries: that the set of numbers $\{0,...,n+m-1\} \setminus \{0,...,n-1\}$ is the same as the set obtained by adding $n$ to everything in $\{0,...,m-1\}$.

@@

\todoi{TG to RK: Feel free to add more examples.}

@@

\textsf{SEPIA}~\cite{DBLP:conf/cade/GransdenWR15} is a powerful system able to
generate
proof scripts from previous \coq proof examples.
Its strength lies in its ability to produce likely sequences
of tactics for solving domain specific goals. It operates by creating a model
for common sequences of tactics for a specific library.
This means that in order to propose the following tactic, only the previously
called tactics
are considered.
%intuition tells us that it should not be the main factor
Our algorithm, on the other hand, relies mainly on the characteristics of the
current goal
to decide
which tactics to apply next. In this way, our learning mechanism has to
rediscover why each
tactic was applied for the current subgoals. It may lack some useful bias for
common sequences
of tactics, but is more reactive to subtle changes. Indeed, it can be trained
on a large library and only tactics relevant to the current subgoal will be
selected.
Concerning the proof search, \textsf{SEPIA}'s %simple
breadth-first search is replaced by MCTS which allows for supervised learning
guidance in the exploration of the search tree.
Finally, \textsf{SEPIA} was evaluated on three chosen parts of the
\coq library demonstrating that it globally outperforms individual \coq
tactics. Here, we demonstrate the competitiveness of our system against
\eprover on the \holfour standard library.
%A recent improvement on this system~\cite{}\todo{Cezary: could not remember
%the
%reference that you gave me here i think it was CICM 2017} includes the
%possibility to
%substitute theorems appearing in tactics by ones belonging to the same
%cluster.
%In this work, we are also able to predict theorems appearing as arguments of
%tactics but we rely only on information from the current goal. Combining the
%two approaches by predicting from the goal and from the original tactic
%arguments is more difficult but is likely to produce more suitable lemmas.

Proof patching: \cite{RingerYLG18}

@@

\section{Conclusion}\label{sec:concl}

We proposed a new proof assistant automation technique which combines
tactic-based proof search with machine learning prediction
Its implementation, \tactictoe, achieves an overall performance of 66.4\%
theorems on 7164 theorems of the \holfour standard library surpassing \eprover
with auto-schedule. Its
effectiveness is especially visible on
theories which use inductive data structures, specialized decision procedures,
and custom built simplification sets.
Thanks to the learning abilities of \tactictoe, the generated proof scripts
usually reveal the high-level structure of the proof. %For these reasons,
We therefore believe that predicting ITP tactics based on the current goal
features is a very reasonable approach to automatically guiding proof search,
and that accurate predictions can be obtained by learning from the knowledge
available in today's large formal proof corpora.

>>>>

%\title{TacticToe: A theorem prover trained from formal proofs}
%\title{TacticToe: Smart guidance in tactical proof search}
%\title{TacticToe: Proof  by Tactic Selection}
\title{TacticToe: Learning to Prove with Tactics}
%\title{TacticToe: Automatic tactic-level proof generation}

@@

\todoi{JU review: Still feels more like an intro than an abstract. Could be
shorter and punchier.}

\todoi{Move part of the abstract to the introduction.}
\begin{abstract}
Formalization of mathematics in interactive theorem provers (ITPs) relies today on many proof automation methods.
Some of these methods are general-purpose and with enough time they can prove any provable conjecture.
Many formalized proof scripts however also take advantage of specialized rules and theory-based strategies.
In the HOL4 ITP system, practically all of them are implemented as \emph{tactics}.

@@

\tableofcontents
\section{Introduction}
Many of the state-of-the-art interactive theorem provers (ITPs) such as
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight},
  \isabelle~\cite{isabelle}
  and \coq~\cite{coq-book} provide high-level parameterizable tactics for constructing proofs.
  Such tactics typically analyze the current proof state (goal and
  assumptions), then apply non-trivial proof transformations that get
  expanded into many basic kernel-level inferences or large parts of a proof term.
%\marginpar{not Coq}
  In this work we develop a tactic-level automation procedure for the \holfour ITP
  which guides the selection of tactics by learning from previous
  proofs.  Instead of relying on translation to first-order automated
  theorem provers (ATPs) as done by the hammer
  systems~\cite{hammers4qed,tgck-cpp15}, our technique
%\marginpar{too many cites?}
%BlanchetteGKKU16,holyhammer,mizAR40
  searches directly for sequences of tactic applications that lead to
  an ITP proof, thus avoiding the translation and proof-reconstruction phases needed by the hammers.

  To do this, we \emph{extract and record} tactic invocations from the ITP
  proofs (Section~\ref{sec:recording}) and
  \emph{build efficient machine learning classifiers} based on such training
  examples (Section~\ref{s:prediction}).  The learned data serves as
  guidance for a \emph{Monte Carlo tree search} that explores different
  proof paths (Section~\ref{sec:proofsearch}). The result, if
  successful, is a verified human-level proof composed of \holfour
  tactics.  The system is evaluated on a large set of theorems originating
  from \holfour (Section~\ref{s:experiments}).

@@

\paragraph{Tactics in \holfour.}
Considering concrete implementation details now, a \emph{tactic} is a function
that takes a goal (or proof obligation) and returns a list of goals (subgoals 
of the original goal) together
with a validation function.
Calling the validation function on a list of \emph{theorems}, corresponding to 
the list of subgoals, results in a theorem corresponding to the original goal.
For example, if the list of subgoals is empty, then calling the validation 
function on the empty list yields the originally goal sequent as a theorem.
In this way, tactics implement the plumbing of backwards, goal-directed, proof 
construction. 
Because validation functions are only important to check the final proof, we 
omit them during the description of proof search algorithm. We note $t(g)$ 
the lists of goals produced by the application of a tactic $t$ to a goal 
$g$.

\subsection{Contributions}
This paper extends the work described in \cite{tgckju-lpar17}.
There we proposed the idea of a tactic-based prover based on supervised 
learning guidance. And we achieved a 39\% success rate on theorems of the 
\holfour standard library by running \tactictoe with a 5 seconds timeout.
The additional contributions and their effect on our system are:

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern
matching constructions, opening of submodules are supported. The different
recording steps (parsing, unfolding and replaying) are regrouped into a single
\sml function.
\item Monte Carlo tree search~\cite{montecarlo} (MCTS) replaces A* as our
search algorithm. The MCTS algorithm gives a more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by predictors for
three kinds of objects: tactics, theorems and lists of goals.
\item The orthogonalization process that eliminates redundant tactics is 
described in a precise manner.
\item A tactic abstraction mechanism is introduced. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover to help \tactictoe during proof search.
\item Evaluation of \tactictoe with a 60 seconds timeout achieving a 66\% 
success rate on the standard library.
\item First evaluation of \tactictoe success rate relative to
the length of the original human proofs.
\item Minimization and prettification of the generated proof improves user 
interaction.
\end{itemize}

@@

\begin{definition}\label{def:prof} (Productive tactic)
The application of a tactic on an open goal is called productive if and only if
all the following conditions are satisfied:
\begin{itemize}
\item It does not fail or timeout. To prevent tactics from looping, we
interrupt tactics after a short amount of time (0.05 seconds).
\item It does not loop, i.e its output does not contain any goals that appears 
in the ancestor of the node of the current goal.
\item It is not a parallel step, i.e its output is not a superset
of a list of goals produced from the same goal.
\end{itemize}

@@

\section{Proof Recording}\label{sec:recording}
Recording proofs in an LCF-style proof assistant can be done at different
levels.
In \holfour all existing approaches relied on modifying the kernel. This was
used
either to export the primitive inference
steps~\cite{Wong95recordingand,DBLP:conf/itp/KumarH12}
or to record dependencies of theorems~\cite{tgck-cpp15}. This was not suitable
for our
purpose of learning proving strategies at the intermediate tactic level. We
therefore
discuss recording proofs in an LCF-style proof assistant, with the focus on
\holfour
in this section.
<
Rather than relying on the underlying programming language, we parse the
original theory file containing proof scripts. This enables us to extract the
code of each tactic.
Working with the string representation of a tactic is better that working with
its value:
\begin{itemize}
\item Equality between two functions is easy to compute from
their string representation and therefore we can avoid
predicting the same tactic over and over again.
\item
The
results of
\tactictoe can be returned in a readable form to the user. In this way, the
user can learn from the feedback, analyze the proof and possibly improve on it.
An other point is that \tactictoe does not need to be installed to run
proof scripts generated by \tactictoe. In this way, the further development of
\tactictoe are not affecting the robustness of \holfour.
\item
It is difficult (probably impossible) to transfer \sml values
between theories as they are not run in the same \holfour session. In contrast,
\sml code can be exported, imported.
\item To produce a tactic value form a code is easy in \sml, which can be
achieved by using a reference to a tactic and update it with the function
\texttt{use}.
\end{itemize}

@@

In Example~\ref{ex:cs3}, we prove a theorem about lists, a domain where \tactictoe excels compared to ATPs.
Given two list $l_1$ and $l_2$ where the length of $l_1$ (noted $p$) is less
than a natural
number $n$, the theorem states updating the $n^{th}$
element of the concatenation of $l_1$ and $l_2$ is the same as
updating the $m^{th}$ element of $l_2$ where $m = n - p$. Again,
\tactictoe's proof is
much more readable starting with applying induction on $l$. It solves the base
case of the induction by rewriting and distinguish cases where $n=0$ or $n>0$
in the inductive hypothesis. It finalizes the proof with a short call to \metis
using the definition of
\texttt{LUPDATE}. Here, \tactictoe's proof is arguably smaller, faster and 
easier to
understand and maintain. Such proofs after an expert review could replace the
their respective original human proofs in the \holfour repository.

@@

In Example~\ref{ex:cs4}, we demonstrate the capabilities of \tactictoe on a 
goal that was not already proved in the \holfour libraries: that the set of 
numbers $\{0,...,n+m-1\} \setminus \{0,...,n-1\}$ is the same as the set 
obtained by adding $n$ to everything in $\{0,...,m-1\}$. In this example, 
\tactictoe uses the simplification set \texttt{ARITH\_ss} to reduce arithmetic 
formulas. This exemplifies an another advantage that \tactictoe has over ATPs, 
mainly its ability to take advantage of user-defined simplification sets.

@@

@@

\textsf{SEPIA}~\cite{DBLP:conf/cade/GransdenWR15} is a powerful system able to
generate
proof scripts from previous \coq proof examples.
Its strength lies in its ability to produce likely sequences
of tactics for solving domain specific goals. It operates by creating a model
for common sequences of tactics for a specific library.
This means that in order to propose the following tactic, only the previously
called tactics
are considered.
%intuition tells us that it should not be the main factor
Our algorithm, on the other hand, relies mainly on the characteristics of the
current goal
to decide
which tactics to apply next. In this way, our learning mechanism has to
rediscover why each
tactic was applied for the current subgoals. It may lack some useful bias for
common sequences
of tactics, but is more reactive to subtle changes. Indeed, it can be trained
on a large library and only tactics relevant to the current subgoal will be
selected.
Concerning the proof search, \textsf{SEPIA}'s %simple
breadth-first search is replaced by MCTS which allows for supervised learning
guidance in the exploration of the search tree.
Finally, \textsf{SEPIA} was evaluated on three chosen parts of the
\coq library demonstrating that it globally outperforms individual \coq
tactics. Here, we demonstrate the competitiveness of our system against
\eprover on the \holfour standard library.
%A recent improvement on this system~\cite{}\todo{Cezary: could not remember
%the
%reference that you gave me here i think it was CICM 2017} includes the
%possibility to
%substitute theorems appearing in tactics by ones belonging to the same
%cluster.
%In this work, we are also able to predict theorems appearing as arguments of
%tactics but we rely only on information from the current goal. Combining the
%two approaches by predicting from the goal and from the original tactic
%arguments is more difficult but is likely to produce more suitable lemmas.
%Proof patching: \cite{RingerYLG18}

@@

\section{Conclusion}\label{sec:concl}
We proposed a new proof assistant automation technique which combines
tactic-based proof search with machine learning prediction
Its implementation, \tactictoe, achieves an overall success rate of 66.4\%
on 7164 theorems of the \holfour standard library, surpassing \eprover
with auto-schedule. Its
effectiveness is especially visible on
theories which use inductive data structures, specialized decision procedures,
and custom built simplification sets.
Thanks to the learning abilities of \tactictoe, the generated proof scripts
usually reveal the high-level structure of the proof. %For these reasons,
We therefore believe that predicting ITP tactics based on the current goal
features is a very reasonable approach to automatically guiding proof search,
and that accurate predictions can be obtained by learning from the knowledge
available in today's large formal proof corpora.
